Beyond the Numbers: Dissecting New York City's Leading Causes of Death Across Demographics

### 2024-03-17 paper/paper.qmd
adding new graphs and tables to quarto

---
title: "Beyond the Numbers: Dissecting New York City's Leading Causes of Death Across Demographics"
subtitle: "A demographics-based analysis of morbditiy in the Big Apple"
author: 
  - Rayan Awad Alim
  - Maria Mangru
  - MD Mubtasim-Fuad 
thanks: "Code and data are available at: https://github.com/RayanAlim/Analysis-of-Morbidity-in-New-York"
date: today
date-format: long
abstract: "We analysed the leading causes of death in New York City, segmented by sex and ethnicity, from 2007 to 2014. Utilizing data derived from NYC death certificates, our research offers insights into public health trends, and the importance of understanding how these causes of death vary across race."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

You can and should cross-reference sections and sub-sections.
We use @citeR and @rohan.

The remainder of this paper is structured as follows.
@sec-data....

@@

## Variables and Data Examination 
The dataset comprises several key variables, including: 
- Year: The year of the recorded death.

@@

## Alternative Datasets and Justification 
While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City.
This specificity provides a more precise tool for understanding and addressing urban health dynamics, making it more relevant for city-specific policy-making and health intervention strategies.

@@

## Summary Statistics and Relationships 
Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City.
We provide summary statistics, including mean death rates and distributions of deaths by cause ethnicity, to provide insights into the health landscape of the city.
For example, preliminary findings indicate significant differences in heart disease mortality between ethnic groups, warranting further investigation.

@@

For our Bayesian regression framework, we employ the stan_glm() function from the rstanarm package, which assumes default normal priors for the regression coefficients as follows:
$\beta_0$ ~ Normal(0,2.5)
$\beta_1$ ~ Normal(0,2.5)

### Model justification
The negative binomial regression was used due to its ability to account for the over-dispersion present in the data. This is a common occurrence with mortality counts due to the heterogeneity of death causes. Through the use of the  parameter, we can model the variance separately from the mean which provides a more flexible and accurate representation of the data. This model also accommodates the count nature of the dependent variable. 
Furthermore, the model’s priors are informed by the normative distributions typically used in Bayesian regression analysis. This enables a degree of regularization that reduces the potential for overfitting.

@@

```
\newpage

@@

>>>>

---
title: "Beyond the Numbers: Dissecting New York City's Leading Causes of Death Across Demographics"
subtitle: "A demographics-based analysis of morbditiy in the Big Apple"
author: 
  - Rayan Awad Alim
  - Maria Mangru
  - MD Mubtasim-Fuad 
thanks: "Code and data are available at: https://github.com/RayanAlim/Analysis-of-Morbidity-in-New-York"
date: today
date-format: long
abstract: "Using Morbidity data for New York City, we analyzed the leading causes of death in New York City, segmented by sex and ethnicity, from 2007 to 2014. Utilizing data derived from NYC death certificates, our research offers insights into public health trends and the importance of understanding how these causes of death vary across races."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries. Despite evidence of the cause of mortality identification inaccuracies (Johns et al., 2013), comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world. The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality. Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation (Cutler et al., 2006, p. 117). Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) (Ruhm, 2000). Another theory suggests that education and ethnicity are the leading causes of lower mortality. For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother. However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children (Pamuk et al., 1998). Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer (Cutler et al., 2006, p. 113) – demonstrating the importance of comparing mortality by cause. From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period. Yet past this period, there is a widespread general divergence. Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality (Cutler et al., 2006, p. 117). Finally, there is a theory that explains differences in mortality by cause due to differing socio-cultural practices by ethnicity and education levels. For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke (Pamuk et al., 1998). While this theory has its merits, the size of its effect is questionable. For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality (Cutler et al., 2006, p. 114).
This overview of mortality causes and the debate on determinants is the motivating context behind this paper. We wish to partially close the gap in understanding between mortality causes and its determinants by analyzing recent data on leading causes of death in New York City, taking a deeper dive into mortality figures by cause and ethnic demographics. We model the leading causes of death by rank and magnitude by ethnicity, hypothesizing that each ethnicity follows a negative binomial distribution. From there, based on relative trends between each ethnicity such as trends in income, education, and health behaviours that supposedly influence mortality, we can infer to a certain extent the validity of existing mortality determinant theories based on the empirical evidence. We find that _______________________________________.
Each section of this paper is organized as follows. Section 2 describes the data and some summary statistics. Section 3 describes our modelling, model hypothesis, and justification. Section 4 presents descriptive statistics and model results. Section 5 is a discussion of the results, including weaknesses, next steps, and conclusive remarks.
Lastly, all analysis was conducted using R and various packages {citeR}

@@

## Variables and Data Examination

The dataset comprises several key variables, including: - Year: The year of the recorded death.

@@

## Alternative Datasets and Justification

While other mortality databases exist, such as the CDC's National Center for Health Statistics, the chosen dataset offers a detailed and localized perspective specific to New York City.
This specificity provides a more precise tool for understanding and addressing urban health dynamics, making it more relevant for city-specific policy-making and health intervention strategies.

@@

## Summary Statistics and Relationships

Initial exploratory data analysis revealed key trends and disparities in mortality rates across different populations within New York City.
We provide summary statistics, including mean death rates and distributions of deaths by cause ethnicity, to provide insights into the health landscape of the city.
For example, preliminary findings indicate significant differences in heart disease mortality between ethnic groups, warranting further investigation.

@@

For our Bayesian regression framework, we employ the stan_glm() function from the rstanarm package, which assumes default normal priors for the regression coefficients as follows: $\beta_0$ \~ Normal(0,2.5) $\beta_1$ \~ Normal(0,2.5)

### Model justification

The negative binomial regression was used due to its ability to account for the over-dispersion present in the data.
This is a common occurrence with mortality counts due to the heterogeneity of death causes.
Through the use of the parameter, we can model the variance separately from the mean which provides a more flexible and accurate representation of the data.
This model also accommodates the count nature of the dependent variable.
Furthermore, the model’s priors are informed by the normative distributions typically used in Bayesian regression analysis.
This enables a degree of regularization that reduces the potential for overfitting.

@@

@@

```{r, echo=FALSE, message=FALSE}

avg_rate_top_causes_by_race <- nyc_data %>%
  group_by(Race.Ethnicity, Leading.Cause) %>%
  summarise(Average_Age_Adjusted_Death_Rate = mean(as.numeric(Age_Adjusted_Death_Rate), na.rm = TRUE)) %>%
  arrange(Race.Ethnicity, desc(Average_Age_Adjusted_Death_Rate)) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5) %>%
  ungroup()

# Rename the columns for display
colnames(avg_rate_top_causes_by_race) <- c("Race/Ethnicity", "Cause of Death", "Average Age-Adjusted Death Rate", "Rank")

# create table 
knitr::kable(avg_rate_top_causes_by_race, caption = "Average Age-Adjusted Death Rates by Top 5 Leading Causes of Death by Race") %>%

  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = F) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
  column_spec(3, width = "3cm")

```{r, echo=FALSE, message=FALSE, fig.cap="Top 5 Causes of Death by Average Age-Adjusted Death Rate.", fig.width=10, fig.height=8, echo=FALSE}

# Prepare the data
avg_rate_top_causes_by_race <- nyc_data %>%
  mutate(Age_Adjusted_Death_Rate = as.numeric(Age_Adjusted_Death_Rate)) %>%
  group_by(Race.Ethnicity, Leading.Cause) %>%
  summarise(Average_Age_Adjusted_Death_Rate = mean(Age_Adjusted_Death_Rate, na.rm = TRUE), .groups = 'drop') %>%
  arrange(Race.Ethnicity, desc(Average_Age_Adjusted_Death_Rate)) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5) %>%
  select(Race.Ethnicity, Leading.Cause, Average_Age_Adjusted_Death_Rate)

# Plotting
gg <- ggplot(avg_rate_top_causes_by_race, aes(x = `Leading.Cause`, y = Average_Age_Adjusted_Death_Rate, fill = `Leading.Cause`)) +
  geom_col() + 
  facet_wrap(~`Race.Ethnicity`, scales = "free_y") + 
  scale_fill_brewer(palette = "Paired") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
        strip.text.x = element_text(size = 6), 
        legend.position = "none") + 
  labs(x = "Top 5 Causes of Death", y = "Average Age-Adjusted Death Rate")

print(gg)
ggsave("avg_rate_top_causes_by_race.png", plot = gg, width = 12, height = 8)

## Model Results

For each population we built a negative binomial and possion regression the model and then use posterior predictive checks to compare the models.
We also compare between the models using the resampling method leave-one-out (LOO) cross-validation (CV).
The results for the models based on each population is as follows:

### White Non-Hispanic Population

Posterior Predictive Check:

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#| label: fig-ppcheckpoissonvsbinomial-white
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of White Non-Hispanic Population in New York City"
#| fig-subcap: ["Negative binomial model", "Poisson model"]

library(magick)

# Load the images
image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_white.png")
image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_white.png")

image1
image2
```

Leave-one-out (LOO) cross-validation (CV):

| Model             | elpd_diff | se_diff |
|-------------------|-----------|---------|
| Negative Binomial | -43.4     | 4.0     |
| Poisson           | 0.0       | 0.0     |

### Black Non Hispanic Population

Posterior Predictive Check:

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#| label: fig-ppcheckpoissonvsbinomial-black
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Black Non-Hispanic Population "
#| fig-subcap: ["Negative binomial model", "Poisson model"]

library(magick)

# Load the images
image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_black.png")
image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_black.png")

image1
image2
```

Leave-one-out (LOO) cross-validation (CV):

| Model             | elpd_diff | se_diff |
|-------------------|-----------|---------|
| Negative Binomial | -37.6     | 3.0     |
| Poisson           | 0.0       | 0.0     |

### Hispanic Population

Posterior Predictive Check:

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#| label: fig-ppcheckpoissonvsbinomial-hispanic
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Hispanic Population in New York City"
#| fig-subcap: ["Negative binomial model", "Poisson model"]

library(magick)

# Load the images
image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_hispanic.png")
image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_hispanic.png")

image1
image2
```

Leave-one-out (LOO) cross-validation (CV):

| Model             | elpd_diff | se_diff |
|-------------------|-----------|---------|
| Negative Binomial | -34.0     | 2.8     |
| Poisson           | 0.0       | 0.0     |

### Asian And Pacific Islanders Population

Posterior Predictive Check:

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#| label: fig-ppcheckpoissonvsbinomial-asian
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for Poisson and negative binomial models in Leading cause of death of Asian And Pacific Islanders Population in New York City"
#| fig-subcap: ["Negative binomial model", "Poisson model"]

library(magick)

# Load the images
image1 <- image_read("../data/analysis_data/Model_Output/pp_check_neg_binom_asian.png")
image2 <- image_read("../data/analysis_data/Model_Output/pp_check_poisson_asian.png")

image1
image2
```

Leave-one-out (LOO) cross-validation (CV):

| Model             | elpd_diff | se_diff |
|-------------------|-----------|---------|
| Negative Binomial | -25.7     | 1.4     |
| Poisson           | 0.0       | 0.0     |
